Did WW2 Produce an Equivalent to a Modern Computer?

### 2017-06-14 Draft 2/Master.tex
Looks like we're done here
Oh wait I'm 3000 words over the limit dearie me

>>>>

\section*{The Limits of Computation}

In our journey, we began with the abstract philosophical concept of computability, knowing nothing beyond roughly what we aim to achieve with it. Defining $\lambda$-Calculus, we were able to see that computation in all its power follows simply from the very natural concept of functions. Defining Turing machines, we saw that this abstract concept is not distant from a physical realisation; and exploring the functionality of Z3, that even relatively simplistic machines can demonstrate computation. It has become clear that the target of computability is not as difficicult to achive as it could appear, and that it was achived before the end of World War 2.

The fact that we have been able to provide a unified definition of logical decision-making along the way is, in my opinion, the most remarkable result of this study. We must consider, however, that there are limits to computation - firstly, there is a set of so called 'undecideable' problems with solutions which cannot be solved algorithmically, including finding integer solutions to Diophantine equations \cite{matiyasevich1993hilbert} and the Halting problem, which would determine whether a given computer program runs indefinitely \cite{turing1936computablenumbers}. Furthermore, when the solution to a problem is probabalistic (e.g. in simulation of quantum processes - see \cite{feynman1982simulating}) as all computation produces the same solution given a problem. As such, despite effective computability being a simple, beautiful and practical theory, it does not achieve Leibniz's aim of a Universal Language; in truth, I believe that the existance of an algorithmic description of everything would be a surprising result. Of course, Mathematics is filled with surprises.

### 2017-06-14 Draft 2/Section 4 - Zuse's Z3.tex
Looks like we're done here
Oh wait I'm 3000 words over the limit dearie me

At this point, we have an almost complete definition of 'computer' - a machine which is able to fully implement the concept of effective computability. This can be done most simply through the implementation of an effectively computable method of computation, such as a Turing Machine. In order to address the question of whether this was achieved during WW1 we must apply this test to the physical instruments created during this period.
\textit{TODO: Colossus}

\textit{TODO: Context of Z3}

\subsection{The Architecture of the Z3}

In this section I shall give a broad functional overview of the Z3. I shall explore primarily the function and usage of the machine, i.e. the aspects that would be visible to the programmer. I shall not go into detail about the operation of the machine itself unless it directly influences the machine's usage. The description of the machine was derived from \cite{rojas1997z3architecture}.

The Z3 was primarily purposed towards performing simple arithmetical calculations - addition, subtraction, multiplication, division, and square root. The operands would be read from memory or from input from a keyboard, and the results would be written to memory or shown on a display. Programs could be written as sequences of 8-bit instructions on punched tape, and would be performed as they are read by the machine.

@@

@@

The instruction set for the Z3 consisted of three types of operation - I/O, memory and arithmetic. These instructions would be performed with respect to two floating-point registers, $R_1$ and $R_2$, which act as the two arguments for any arithmetical operation, and could be written to and read from by the non-arithmetical operations.

@@

The arithmetical operations perform as follows:
\begin{gather*}
\begin{aligned}
&\mathrm{Ls_1}:\quad	&R_1 &= R_1 \cdot R_2 \\
&\mathrm{Ls_2}:\quad	&R_1 &= R_1 \div R_2 \\
&\mathrm{Lm}:\quad	&R_1 &= R_1 + R_2 \\
&\mathrm{Li}:\quad	&R_1 &= R_1 - R_2 \\
&\mathrm{Lw}:\quad	&R_1 &= \sqrt{R_1} \\
\end{aligned}
\end{gather*}
For each operation, the operation is performed on $R_1$ and $R_2$, and the result is written to $R_1$. 
\subsection{Universality in the Z3}

@@

At first glance, it is unclear that the Z3 can be made into a universal machine due to the fact that it allows for neither conditional branching\footnote{The ability to perform different operations depending on the state of a stored value} nor indirect addressing\footnote{The ability to refer to values stored in memory when the address of that memory is not specifically referred to in the program but is rather obtained from the value of some variable currently in memory}. However, through use of the tools available to the programmer these features can be emulated, and a Turing Machine can be simulated. This uses many tecniques explained in \cite{rojas1998z3universal}.

@@

Given a positive integer variable $z$ on which the outcome of the statement depends and the set of operations to be performed depending on the value of $z$, $P_1, ... , P_n$, we can produce a sequence of commands that will allow only the operation $P_z$ to perform. Due to the nature of the Z3 requiring a full sequence of commands to be executed without exception, every one of these operations will be evaluated; however, only operation $P_z$ will be allowed to actually affect memory.

@@

Using this method to emulate switch statements, we can produce an equivalent to an indirect addressing operation. This operation is of form $y = (x)$, which sets $n$ to the value at address location $x$, where $x$ is an integer $n_a \leq x \leq n_b$, $n_a$ and $n_b$ being integer constants representing the limits of the region of memory that can be addressed by this function. $x$ could potentially be the output from a previous arithmetical operation. Relying on the fact that this region is finite (and furthermore that the tape length is also finite), this can be done using a switch statement with a case for every address $z$ between $n_a$ and $n_b$ as shown.
\begin{gather*}
\begin{aligned}
&\bm{\mathrm{switch}}\ (x)\ \{	\\
&\qquad ... ,					\\
&\qquad z:\	\mathrm{Pr}\ z;\ \mathrm{Ps}\ a			\\
&\qquad ... 					\\
&\}								\\
\end{aligned}
\end{gather*}
At this point, it becomes very clear the extent to which the Z3 is not suited to universal computation. When every single applicable memory address must be specified individually, with five operations per address to implement the case, each operation consisting of four instructions, it is obvious that the Z3 is in no way a \textit{practical} universal machine.

To emulate a Turing machine using these new tools, one considers the tape to be the region of memory between $n_a$ and $n_b$, and 
\end{document}

>>>>

At this point, we have a complete definition of `computer' - a device which is able to fully implement the concept of effective computability. This can be done most simply through the emulation of an effectively computable method of computation, such as a Turing Machine. In order to address the question of whether this was achieved during WW1 we must apply this test to the physical instruments created during this period. The

Z3, completed in Berlin in 1941 by Konrad Zuse, was arguably the first programmable computer ever built. A far faster and more reliable version of its two predecessors, Z1 and Z2, Z3 was advanced because of its ability to carry out long sequences of calculations, written as instructions on punched tape, with the potential to store values in memory. While still very simplistic, and not directly intended for universal computation, Z3 is structurally far closer to achieving this aim than other machines of the time; Hence, in this section I shall attempt to implement a Universal machine in Z3 in order to determine its nature as a computing machine.

\subsection{The Architecture of Z3}

In this section I shall give a broad functional overview of Z3. I shall explore primarily the function and usage of the machine, i.e. the aspects that would be visible to the programmer. I shall not go into detail about the operation of the machine itself unless it directly influences the machine's usage. The description of the machine was derived from \cite{rojas1997z3architecture}.

Z3 was primarily purposed towards performing simple arithmetical calculations - addition, subtraction, multiplication, division, and square root. The operands would be read from memory or from input from a keyboard, and the results would be written to memory or shown on a display. Programs could be written as sequences of 8-bit instructions on punched tape, and would be performed as they are read by the machine.

@@

\textit{Basically it can store decimals, and these integers are referred to by a numeric 'address'. for the moment we'll use letters instead of numbers because we're cool like that}

@@

The instruction set for Z3 consisted of three types of operation - I/O, memory and arithmetic. These instructions would be performed with respect to two floating-point registers, $R_1$ and $R_2$, which act as the two arguments for any arithmetical operation, and could be written to and read from by the non-arithmetical operations.

@@

The arithmetical operations perform as follows:
\begin{gather*}
\begin{aligned}
&\mathrm{Ls_1}:\quad	&R_1 &= R_1 \cdot R_2 \\
&\mathrm{Ls_2}:\quad	&R_1 &= R_1 \div R_2 \\
&\mathrm{Lm}:\quad	&R_1 &= R_1 + R_2 \\
&\mathrm{Li}:\quad	&R_1 &= R_1 - R_2 \\
&\mathrm{Lw}:\quad	&R_1 &= \sqrt{R_1} \\
\end{aligned}
\end{gather*}
For each operation, the operation is performed on $R_1$ and $R_2$, and the result is written to $R_1$. 
\subsection{Universality in Z3}

@@

At first glance, it is unclear that Z3 can be made into a universal machine due to the fact that it allows for neither conditional branching\footnote{The ability to perform different operations depending on the state of a stored value} nor indirect addressing\footnote{The ability to refer to values stored in memory when the address of that memory is not specifically referred to in the program but is rather obtained from the value of some variable currently in memory}. However, through use of the tools available to the programmer these features can be emulated, and a Turing Machine can be simulated. This uses many tecniques explained in \cite{rojas1998z3universal}.

@@

Given a positive integer variable $z$ on which the outcome of the statement depends and the set of operations to be performed depending on the value of $z$, $P_1, ... , P_n$, we can produce a sequence of commands that will allow only the operation $P_z$ to perform. Due to the nature of Z3 requiring a full sequence of commands to be executed without exception, every one of these operations will be evaluated; however, only operation $P_z$ will be allowed to actually affect memory.

@@

Using this method to emulate switch statements, we can produce an equivalent to an indirect addressing operation. This operation is of form $y = (x)$, which sets $y$ to the value at address location $x$, where $x$ is an integer $n_a \leq x \leq n_b$, $n_a$ and $n_b$ being integer constants representing the limits of the region of memory that can be addressed by this function. $x$ could potentially be the output from a previous arithmetical operation, which could be read from memory. Relying on the fact that this region is finite (and furthermore that the tape length is also finite), this can be done most simply with a switch statement with a case for every address $z$ between $n_a$ and $n_b$ as shown.
\begin{gather*}
\begin{aligned}
&\bm{\mathrm{switch}}\ (x)\ \{              	\\
&\qquad ... ,                                   \\
&\qquad z:\	\mathrm{Pr}\ z;\ \mathrm{Ps}\ a,    \\
&\qquad ... 					                \\
&\}							                    \\
\end{aligned}
\end{gather*}
These cases are enumerated within the switch for all $n_a \leq z \leq n_b$; hence, when the statement is executed, the only case where the contents of address $z$ will be written to $a$ is the case where $x=z$; hence, $a$ will be set to the contents of address $x$.
At this point, it becomes very clear the extent to which  is not suited to universal computation. When every single applicable memory address must be specified individually, with five operations per address to implement the case, each operation consisting of four instructions, it is obvious that Z3 is in no way a \textit{practical} universal machine.

\subsection{Z3 as a Turing Machine}

With these new methods implemented in Z3, construction of a Universal machine seems trivial. As shown in \cite{harel1980folk}, a Turing machine can be produced using only a single continuous loop and the functionality of Z3. The ends of the instruction tape can be attached to each other as they are run through the machine, allowing for a continuous loop. The tape is represented as the memory region between $n_a$ and $n_b$, and the cursor index is stored as an integer that can be moved through incrementation or decrementation. The tape can be read through indirect addressing, and the transition can be derived from the state and read symbol through conditional branching.

The result of this section, that arithmetic alone can produce universality, is surprising and counterintuitive. However, the fundamental concept behind this is that within each arithmetical operation are conditional branches that the processor performs; Turing-completeness is embedded within Z3's operation, albeit invisible to the programmer. These tools are intended to lift this branching to the surface level so they can control the program's flow.

A clear criticism of this techique is that it slows down computation enormously. Running at roughly 5.33Hz with 3 cycles for addition, 4-5 for subtraction and 16 for multiplication (from \cite{rojas1997z3architecture}), to perform each case statement within a switch Z3 would take up to $\sim$7.5 seconds on top of the time for the operation itself; for the indirect addressing operation alone, when this occurs for every single addressable memory slot it is clear that for a single iteration of the Turing machine Z3 would take an unfeasably long time. From a purely theoretical point of view this is irrelevant - it is clear that this particular programming of Z3 is useless in the same way that using a Turing Machine to solve real-world problems would be totally impractical.

Therefore, we can say that from a totally abstract, totally theoretical perspective, World War 2 did in fact produce an equivalent to a modern computer. However, like many of the computers of this period, while effectively equivalent to a modern computer, Z3 was entirely unlike anything one would encounter today and as such would not be suited to perform anything beyond its simplistic intended use.

The major breakthrough in computation which allowed for the first clearly universal machines was the implementation of a Von Neumann Architecture, which allowed computers to run programs from memory\footnote{Of course, because Turing Machines are themselves Universal machines (see \cite{turing1936computablenumbers}), this was theoretically achieved with Z3; Von Neumann machines were oriented towards performing in such a way.}. The first machine to implement this was the Manchester Small Scale Experimental Machine (SSEM), or `Baby' \cite{napperWEB17baby}, which stored data using the Williams-Kilburn Tube, a type of cathode ray tube. Its successor, the Manchester Mark 1 contained a more advanced instruction set, and was significantly more powerful.

\end{document}
